Cereal Nutrition by Mark Robert Dods
========================================================
  
```{r echo=FALSE, message=FALSE, warning=FALSE}

#install.packages('dplyr', dependencies = TRUE)
#install.packages('ggplot2', dependencies = TRUE)
#install.packages('gridExtra', dependencies = TRUE)
#install.packages('GGally', dependencies = TRUE)
#install.packages('memisc', dependencies = TRUE)
#install.packages('psych', dependencies = TRUE)
#install.packages('knitr', repos = "http://cran.us.r-project.org")


library(ggplot2)
library(dplyr)
library(gridExtra)
library(GGally)
library(memisc)
library(psych)
library(knitr)

setwd("C:/r")

cereals <- read.csv('cereal.csv', header = T,check.names = F)
cereals <- cereals[!(cereals$mfr=="A"),]
cereals <- plyr::rename(cereals, c("mfr" = "manufacturer"))
cereals$manufacturer <- plyr::mapvalues(cereals$manufacturer, 
  from=c("G", "K","N", "P", "Q","R"), to=c("General Mills", "Kelloggs",
  "Nabisco", "Post", "Quaker Oats","Ralston Purina"))
cereals <- mutate(cereals, SugarsToFiber = sugars/fiber)
cereals <- mutate(cereals, SugarsToFat = sugars/fat)
cereals <- mutate(cereals, WeightInGrams = weight * 28.3495)
cereals <- mutate(cereals, PercFat = fat / WeightInGrams)
cereals <- mutate(cereals, PercSodium = sodium / WeightInGrams)
cereals <- mutate(cereals, PercSugar = sugars / WeightInGrams)
cereals <- mutate(cereals, PercFiber = fiber / WeightInGrams)

```

### Introduction:  

I chose the '80 cereals' dataset that I found on Kaggle
(https://www.kaggle.com/crawford/80-cereals).  I chose it because it had 16 
different variables (13 dependent, 3 independent) associated with each cereal 
which I could compare to see which ones are correlated. I focused mainly on 
sugar, fat, sodium, and fiber, and investigated 1) which manufacturer makes 
the most healthy or unhealthy cereal, and 2) which variable most affects the 
overall rating given to the cereal. I then created a model to predict rating 
based on the sugar, fat, sodium, and fiber content of a particular cereal.

### Fields in the dataset:
   
* Name: Name of cereal  
* manufacturer: manufacturer of cereal  
    + A = American Home Food Products   
    + G = General Mills  
    + K = Kelloggs  
    + N = Nabisco  
    + P = Post  
    + Q = Quaker Oats  
    + R = Ralston Purina  
* type:  
    + cold  
    + hot  
* calories: calories per serving  
* protein: grams of protein  
* fat: grams of fat  
* sodium: milligrams of sodium  
* fiber: grams of dietary fiber  
* carbo: grams of complex carbohydrates  
* sugars: grams of sugars  
* potass: milligrams of potassium  
* vitamins: vitamins and minerals - 0, 25, or 100, indicating the typical 
percentage of FDA recommended  
* shelf: display shelf  
* weight: weight in ounces of one serving  
* cups: number of cups in one serving  
* rating: a rating of the cereals   

## Univariate Plots Section

Number of rows in dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}
nrow(cereals) #find number of rows in dataset

```

Summary of data type for each variable in dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(cereals)  #summary of data type held by each variable in dataset

```

Summary stats for all variables in dataset

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(cereals) #summary of all variables in dataset

```

Number of cereals in dataset by manufacturer


```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(cereals$manufacturer) #shows number of cereals by manufacturer

#by(cereals$sugars,cereals$manufacturer,summary) 
#by(cereals$fat,cereals$manufacturer,summary) 
#by(cereals$sodium,cereals$manufacturer,summary) 
#by(cereals$fiber,cereals$manufacturer,summary) 


```

Distributions of selected nutritional variables

Note:  All statistical outliers have been included as they are valid values
(as opposed to abnormal ones) within the distributions and are significant to 
the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals, aes(sugars)) + geom_histogram(binwidth=0.5)  +
  scale_x_continuous(limits = c(0,15),breaks = seq(0,15,1))



```

The above plot shows the Distribution of Sugars.  There is a spike at 3g so it 
would appear that this is a target sugar content for many manufacturers.  I 
edited the x axis values to exclude a value of -1 which was obviously an error 
in data entry as it is an impossible value.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals, aes(fat)) + geom_histogram(binwidth=0.5)  +
  scale_x_continuous(breaks = seq(0,5,1))

```

The above plot shows the Distribution of Fat, and we can see that the majority
of cereals contain either 1 or 2 grams of fat.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals, aes(sodium)) + geom_histogram(binwidth=2.75)  +
  scale_x_continuous(breaks = seq(0,325,25))

```

The above plot shows the Distribution of Sodium. Sodium content varies a lot, 
with 0 and 200mg being spikes, but many other values exist, all the way from 0
to 320.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals, aes(fiber)) + geom_histogram(binwidth=0.2)  +
  scale_x_continuous(breaks = seq(0,20,1))

```

The above plot shows the Distribution of Fiber. We can see that the majority
of cereals have very low fiber, with 0 being the most common amount, and we 
can see that the majority of cereals have 3g or less of fiber.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals, aes(rating)) + geom_histogram(binwidth=2)  +
  scale_x_continuous(breaks = seq(0,100,5))

```

The above plot shows the Distribution of Ratings. We can see that 40 is the most
commmon rating, but it's hard to draw conclusions from the frequencies of the 
ratings, it would be good to have more data to draw from.

```{r echo=FALSE, message=FALSE, warning=FALSE}


```
## Univariate Analysis

### What is the structure of your dataset?

The structure of my dataset is 76 rows with 16 original columns, and 
7 more that I added.

### What is/are the main feature(s) of interest in your dataset?

I am focusing on sugar, fat, sodium, and fiber and specifically on the
relationship between said variables and the overall rating assigned to them.

### What other features in the dataset do you think will help support your /
investigation into your feature(s) of interest?
   
It helps that the dataset is complete and does not have any NA values.

### Did you create any new variables from existing variables in the dataset?

Yes, 7, as follows  
*
cereals <- mutate(cereals, SugarsToFiber = sugars/fiber)  
cereals <- mutate(cereals, SugarsToFat = sugars/fat)  
cereals <- mutate(cereals, WeightInGrams = weight * 28.3495)  
cereals <- mutate(cereals, PercFat = fat / WeightInGrams)  
cereals <- mutate(cereals, PercSodium = sodium / WeightInGrams)  
cereals <- mutate(cereals, PercSugar = sugars / WeightInGrams)  
cereals <- mutate(cereals, PercFiber = fiber / WeightInGrams)*

### Of the features you investigated, were there any unusual distributions?  
### Did you perform any operations on the data to tidy, adjust, or change the 
### form of the data? If so, why did you do this?
  
There weren't any distributions I would characterize as unusual, but I think
the distribution of Sodium was intersting in that it was the only one that 
had a distribution that was widely distributed and more random than the others.

I did not need to adjust or change the form of the data as it was already in a
tidy form.



## Bivariate Plots Section

In this section I want to look at the selected nutritional variables by 
manufacturer, along with their distribution by manufacturer.  I think it also 
might be interesting to look at the relationships between fat and sugar as I 
have heard that manufacturers try to raise one when they lower the other, and 
vice versa.  I also want to look at the relationship between sugar and fiber as
I know that some of the harmful of effects of consuming sugar are mitigated when
the sugar is accompanied by an equal or great amount of fiber.

These plots below show nutritional variables by manufacturer

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(y = sugars,x = manufacturer, fill = manufacturer))  + 
  stat_summary(fun.y="mean", geom="bar") + 
  xlab("Manufacturer") + 
  ylab("Sugar Content") + 
  ggtitle("Average Sugar Content by Manufacturer") +
  scale_x_discrete(label=abbreviate) + 
  theme(axis.text.x = element_text(angle = 15, hjust = 1)) + guides(fill=FALSE)

```

The above plot shows that Nabisco cereals have the lowest average sugar content, 
while Post cereals have the highest.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(y = fat,x = manufacturer, fill = manufacturer))  + 
  stat_summary(fun.y="mean", geom="bar") + 
  xlab("Manufacturer") + 
  ylab("Fat Content") + 
  ggtitle("Average Fat Content by Manufacturer") +
  scale_x_discrete(label=abbreviate) + 
  theme(axis.text.x = element_text(angle = 15, hjust = 1)) + guides(fill=FALSE)

```
 
The above plot shows that Nabisco cereals have the lowest average fat content, 
while Quaker Oats have the highest.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(y = sodium,x = manufacturer, fill = manufacturer))  + 
  stat_summary(fun.y="mean", geom="bar") + 
  xlab("Manufacturer") + 
  ylab("Sodium Content") + 
  ggtitle("Average Sodium Content by Manufacturer") +
  scale_x_discrete(label=abbreviate) + 
  theme(axis.text.x = element_text(angle = 15, hjust = 1)) + guides(fill=FALSE)

```

The above plot shows that Nabisco cereals have the lowest average sodium 
content, while General Mills cereals have the highest.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(y = fiber,x = manufacturer, fill = manufacturer))  + 
  stat_summary(fun.y="mean", geom="bar") + 
  xlab("Manufacturer") + 
  ylab("Fiber Content") + 
  ggtitle("Average Fiber Content by Manufacturer") +
  scale_x_discrete(label=abbreviate) + 
  theme(axis.text.x = element_text(angle = 15, hjust = 1)) + guides(fill=FALSE)


```

The above plot shows that Nabisco cereals have the highest fiber content, while 
General Mills cereals have the lowest.

```{r echo=FALSE, message=FALSE, warning=FALSE}
  
ggplot(cereals,aes(y = rating,x = manufacturer, fill = manufacturer))  + 
  stat_summary(fun.y="mean", geom="bar") + 
  xlab("Manufacturer") + 
  ylab("Rating") + 
  ggtitle("Rating by Manufacturer") +
  scale_x_discrete(label=abbreviate) + 
  theme(axis.text.x = element_text(angle = 15, hjust = 1)) + guides(fill=FALSE)


```

The above plot shows that Nabisco has the highest average cereal rating, while 
the rest are quite even, although General Mills has the lowest average rating.


The box plots below show distribution of data for selected nutritional variables
by manufacturer

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=manufacturer,y=sugars)) + geom_boxplot() + 
  xlab("Manufacturer") + ylab("Sugar Content") + 
  ggtitle("Sugar Content by Manufacturer") + 
  scale_x_discrete(label=abbreviate) 

```

The above plot shows that Nabisco cereals have the smallest distribution of 
sugar content, while Kellogs and Quaker Oats have the largest. It is interesting 
that there are no outliers anywhere.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=manufacturer,y=fat)) + geom_boxplot() + 
  xlab("Manufacturer") + ylab("Fat Content") + 
  ggtitle("Fat Content by Manufacturer") + 
  scale_x_discrete(label=abbreviate) 

```

The above plot shows that all Nabisco cereals have 0 fat content, while Ralston 
Purina has the largest distribution. There are outliers, with Quaker Oats having
the most extreme one.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=manufacturer,y=sodium)) + geom_boxplot() + 
  xlab("Manufacturer") + ylab("Sodium Content") + 
  ggtitle("Sodium Content by Manufacturer") + 
  scale_x_discrete(label=abbreviate) 

```

The above plot shows that General Mills cereals have the smallest distribution 
of sodium content (but with 2 outliers) while Quaker Oats has the largest 
distribution. Both Ralston Purina and Post have an outlier that contains less 
sodium than their average.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=manufacturer,y=fiber)) + geom_boxplot() + 
  xlab("Manufacturer") + ylab("Fiber Content") + 
  ggtitle("Fiber Content by Manufacturer") + 
  scale_x_discrete(label=abbreviate) 

```

The above plot shows that Nabisco cereals have the smallest distribution, all 
have relatively high fiber content (with one outlier above and one below), while
Post cereals have the largest distribution.  Kellogs also has 2 high outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=manufacturer,y=rating)) + geom_boxplot() + 
  xlab("Manufacturer") + ylab("Rating") + 
  ggtitle("Rating by Manufacturer") + 
  scale_x_discrete(label=abbreviate) 

```

The above plot shows that Nabisco has the highest cereal rating and the lowest 
distribution, which means they don't have a single cereal with bad ratings. 
Quaker Oats had the largest distribution, and Kellogs had one extremly high 
outlier.


Correlation matrices for selected variables
```{r echo=FALSE, message=FALSE, warning=FALSE}

pairs.panels(cereals[c("fat","sodium","fiber","sugars","rating")],76) 

```

The above correlation matrix shows correlation indices for rating vs. Fat, 
Sodium, Fiber, and Sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE}

pairs.panels(cereals[c("rating","PercFat", "PercSodium", "PercFiber", 
                       "PercSugar")],76)

```

The above correlation matrix shows correlation indices for rating vs. PercFat, 
PercSodium, PercFiber, and PercSugar. Where Perc is the percentage of content to
total weight.  

  
```{r echo=FALSE, message=FALSE, warning=FALSE}

p1 = ggplot(cereals,aes(x = PercSugar, y = rating,colour = rating)) + 
  geom_jitter() + geom_smooth(method='lm') + scale_x_log10() +
  ggtitle("Rating vs PercSugar vs Rating") 

p2 = ggplot(cereals,aes(y = rating, x = PercFat,colour = rating)) + 
  geom_jitter(width = 0.25)  +
  ggtitle("Rating vs PercFat vs Rating") 

p3 = ggplot(cereals,aes(y = rating, x = PercSodium,colour = rating)) + 
  geom_jitter(width = 0.25) +
  ggtitle("Rating vs PercSodium vs Rating") 

p4 = ggplot(cereals,aes(y = rating, x = PercFiber,colour = rating)) + 
  geom_jitter() + 
  geom_smooth(method='lm') + scale_x_log10() +
  ggtitle("Rating vs PercFiber vs Rating")

grid.arrange(p1,p2,p3,p4)

```

## Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
### investigation. How did the feature(s) of interest vary with other features 
### in the dataset?

I found that there was no significant correlation between either Sugar and Fat 
(weak) or Sugar and Fiber (very weak). I then looked at the correlation between 
rating and percentage of the selected nutrirional values (since percentage is a 
more accurate measure then the number of grams/mg since the serving sizes can be
different). I found that there is a strong correlation between the rating and 
both percentage of sugar and percentage of fiber.  Although the correlation 
between rating and percentage of fat and percentage of sodium is lower, I will 
still use them to contruct my model to predict the rating of a cereal based on 
its nutrition values.
  
### Did you observe any interesting relationships between the other features
### (not the main feature(s) of interest)?

Yes, a strong correlation that I wasn´t studying was the correlation betweeen 
potassium and fiber (0.91)

  
### What was the strongest relationship you found?

The strongest correlation value found in the matrix is -0.92 for calories and 
fiber.
  
  
## Multivariate Plots Section
  

These plots show the ratio of 1) Sugar to Fat and 2) Sugar to  # Fiber, first 
all on one plot, then faceted by manufacturer.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=sugars,y=fat,color = manufacturer)) + 
  geom_jitter(size = 2) +
  geom_vline(xintercept = 7.5) + geom_hline(yintercept = 2.5) 


ggplot(cereals,aes(x=sugars,y=fat,color = manufacturer)) + 
  geom_jitter(size = 2) +
  geom_vline(xintercept = 7.5) + geom_hline(yintercept = 2.5) +
  facet_wrap(~manufacturer)

```

The healthiest quadrant for the above 2 plots is the lower left one, and the  
unhealthiest, the upper right one.

```{r echo=FALSE, message=FALSE, warning=FALSE}
  
ggplot(cereals,aes(x=sugars,y=fiber,color = manufacturer)) + 
  geom_jitter(size = 2) +
  geom_vline(xintercept = 7.5) + geom_hline(yintercept = 7) 

ggplot(cereals,aes(x=sugars,y=fiber,color = manufacturer)) + 
  geom_jitter(size = 2) +
  geom_vline(xintercept = 7.5) + geom_hline(yintercept = 7) + 
  facet_wrap(~manufacturer)

```

The healthiest quadrant for the above 2 plots is the upper left one, and the 
unhealthiest, the lower right one.


```{r echo=FALSE, message=FALSE, warning=FALSE}

```

Correlation Indices for Sugar to Fiber

```{r echo=FALSE, message=FALSE, warning=FALSE}

Sugars2FiberRatio <- cereals$PercSugar /cereals$PercFiber
Sugars2FiberRatio[Sugars2FiberRatio == NaN] <- 0
Sugars2FiberRatio[Sugars2FiberRatio == Inf] <- 10
rating2sugars2fiber <- cor.test(cereals$rating,Sugars2FiberRatio)
rating2sugars2fiber

ggplot(cereals,aes(x=rating, y = PercSugar / PercFiber)) + 
geom_jitter() + geom_smooth(method ="lm") +
   ggtitle("Rating vs PercSugar / PercFiber")  

```

So the ratio of percentage sugars to percentage fiber is a decent predictor of 
rating, but for my model I will add percentage sodium and percentage fat as 
well.

```{r echo=FALSE, message=FALSE, warning=FALSE}

fit <- lm(rating ~PercSugar + PercFiber + PercSodium + PercFat, data=cereals)
summary(fit)

m1 <- lm(I(rating) ~ I(PercSugar), data = cereals)
m2 <- update(m1, ~ . + PercFiber)
m3 <- update(m2, ~ . + PercSodium)
m4 <- update(m3, ~ . + PercFat)
mtable(m1, m2, m3, m4)

```

## Multivariate Analysis

I observed the rating and how it was affected by the relationship between the 
percentages of sugar, fat, sodium, and fiber.

### Were there any interesting or surprising interactions between features?

I didn´t find any surprising interactions, the effects on the rating are what I 
would have predicted.

### OPTIONAL: Did you create any models with your dataset? 

Yes, I made a model to predict rating based on PercSugar,PercFiber,PercSodium, 
and PercFat.the model was very accurate, with an R-squared value of 0.931 with 
all 4 varibles included.

------

## Final Plots and Summary

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE}
  
  ggplot(cereals,aes(x=manufacturer,y=rating)) + geom_boxplot() + 
  xlab("Manufacturer") + ylab("Rating") + 
  ggtitle("Rating by Manufacturer") + 
  scale_x_discrete(label=abbreviate) 

```

### Description One

This box plot plot shows how much higher Nabisco is than all the other 
manufacturers. You can see that other manufacturer´s highest rated cereals are 
only comparable with Nabisco´s lowest rated ones.

### Plot Two

```{r echo=FALSE, message=FALSE, warning=FALSE}

p1 = ggplot(cereals,aes(x = PercSugar, y = rating,colour = rating)) + 
  geom_jitter() + geom_smooth(method='lm') + scale_x_log10() +
  ggtitle("Rating vs PercSugar vs Rating") 

p2 = ggplot(cereals,aes(y = rating, x = PercFat,colour = rating)) + 
  geom_jitter(width = 0.25)  + ggtitle("Rating vs PercFat vs Rating") 

p3 = ggplot(cereals,aes(y = rating, x = PercSodium,colour = rating)) + 
  geom_jitter(width = 0.25) + ggtitle("Rating vs PercSodium vs Rating") 

p4 = ggplot(cereals,aes(y = rating, x = PercFiber,colour = rating)) + 
  geom_jitter() + geom_smooth(method='lm') + scale_x_log10() +
  ggtitle("Rating vs PercFiber vs Rating") 

grid.arrange(p1,p2,p3,p4)

```

### Description Two

I like these 4 plots together as you can see how their respective correlation 
indices look like when plotted.  It is obvious that PercSugar is closely 
correlated to rating, while PercFat looks almost randomly distributed.

### Plot Three

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(cereals,aes(x=rating, y = PercSugar / PercFiber)) + 
geom_jitter() + geom_smooth(method ="lm") +
   ggtitle("Rating vs PercSugar / PercFiber")  

```

### Description Three

Without trying to find a more complex model, I thought that the ratio of 
Percentage of Sugar to Percentage of Fiber correlated well with the rating.  At 
the top of the plot, we find all the cereals who had 0 fiber, this is a drawback
of using a ratio to plot.

------
  
## Reflection

My overall conclusion is 1) that Nabisco makes the most nutritious cereals, by 
far, and 2) that the rating for each cereal can be closely approximated by using
a predictive model of PercSugar + PercFat +PercSodium + PercFiber.
  
Some of the challenges I encountered were: not knowing what year the study was 
from, not knowing where or how the rating came from,  dealing with a relatively 
small and incomplete dataset, and deciding what the most relevant columns were 
to cereal nutrition.  Ideally, in the future, the dataset would include many 
more cereals so that there is more data to base conclusions off of.  It would be
interesting in the future to examine the relationship between ratings and 
manufacturer, to see if there is any favoritism involved, or if the ratings are 
generated from a fixed formula.

## Resources used

https://www.kaggle.com/crawford/80-cereals

https://stackoverflow.com/questions/18882206/change-outline-and-fill-colors-of
-histogram-with-qplot

https://stackoverflow.com/questions/40711980/r-ggplot2-fit-curve-to-scatter-plot

https://stackoverflow.com/questions/22282531/how-to-compute-correlations-between
-all-columns-in-r-and-detect-highly-correlate

https://stackoverflow.com/questions/8214303/conditional-replacement-of-values-in
-a-data-frame

https://stackoverflow.com/questions/36590230/how-to-remove-rows-with-inf-from-a-
dataframe-in-r/36590347










